\jaxcosmo: End-to-End Differentiable and GPU Accelerated Cosmology Library

### 2022-03-04 paper.tex
relecture de la discussion

The results of the evolution of the 2D parameter contours for some cosmological parameters compared to the default NUTS sampling (Sec.~\ref{sec-NUTS}) are displayed in figures \ref{fig_svi_nuts_1}, \ref{fig_svi_nuts_2}, \ref{fig_svi_nuts_3}. The agreement is quite good even with a rather low number of samples (ie. 200) as shown on figure \ref{fig_svi_nuts_1} and even for the highly correlated $(b_i)_{i:1,\dots,5}$ bias parameters (Fig.~\ref{fig_svi_nuts_4}). It is noticeable that running NUTS with the same settings  but without the SVI Neural Transport phase has demonstrated a worse behaviour.  Results in terms of sampling efficiency are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1} showing that the SVI followed by Neutral Transprt gives high efficiency as well without the need by the user to setup a particular mass matrix structuring (ie.\texttt{dense} matrix option). 
%
\subsection{Discussion}
%
The question that one may naturally arises are what is the HMC/NUTS gain compared to the well known standard \texttt{Cobaya} sampler? 
In one hand, one may argue considering the elapse time of execution that the number of efficient sample per model call is a right metric to consider: 
\begin{equation}
    \eta = \frac{n_{eff}}{\#\ of\ calls} = \frac{N_s \times \varepsilon}{N_s \times n_{steps}} = \frac{\varepsilon}{n_{steps}}
\end{equation}
with $N_s$ the total number of samples, $\varepsilon$ the sampler efficiency and $n_{steps}$ the number of steps (calls) per sample generation. Then, for \texttt{cobaya} one roughly gets $\varepsilon\approx 1\%$ with $n_{steps}=1$ while for the NUTS sampler $\varepsilon\approx 50\%$ but at the expense of $n_{steps}=2^5$ or more. At the end of the day, the $\eta$ values for both methods are comparable (ie. about $1\%$). So, at first sight considering the elapse time that there might be no gain to use HMC/NUTS sampler compared to standard MCMC sampler.

On the other hand, the above metric is one aspect of the sample generation. One should have in mind that the low sampling efficiency of standard MCMC sampler requires mobilizing a large amount of resources to produce a sufficient large sample batch in a reasonable time scale, ie. several days on dedicated infrastructure. Moreover, the user is faced to a dilemma to perform or not a statistical sample selection post-processing as \textit{thinning} beyond the \textit{burn-in removal} which is a rather delicate procedure and time consuming too, needing how-know to be conducted correctly \citep{doi:10.1146/annurev-statistics-040220-091727, 2015arXiv151007727O}. So, the authors think that it is worth considering methods that by construction are more efficient. Moreover, using SVI with neural reparametrisation offers an effective way to undertake a progressive validation of the model with rather modest sample size (eg. 200 samples) by producing good quality marginal contours in few hours. In practice, this validation phase can save time before producing sizeable batch for final analysis. The authors have not investigated very high dimensional ($O(10^2)$ parameters) as well as multi-modal analysis but first HMC/NUTS sampling exploit the geometry of the typical set of the posterior distribution contrary to the standard random walk of Metropolis-Hasting sampling and further more using reparametrisation (eg. \cite{2019arXiv190303704H}) one can adapt to poor geometry cases.
%
\begin{table}[htb]
\caption{Results of the relative effective sample size (aka ESS, "bulk") in percent computed by the \texttt{Arviz} library \citep{arviz_2019}.
%{\color{red} 
%(0) "F.L run" are computed from NUTS sampler with 16 %parallel chains of 1,000 samples each and 200 warm-up %and \texttt{max\_tree\_depth=7} (see text) The numbers %are from \texttt{n\_eff/16,000};
%}
(1): results from NUTS sampler with 16 chains of 1,000 samples each and 200 warm-up and \texttt{max\_tree\_depth=7} (Sec.~\ref{sec-NUTS}); (2) are the results of SVI Multivariate Normal followed by NUTS and Neural Transform  with 10 chains of 1,000 samples each and 200 warm-up (Sec.~\ref{sec-Neural-Reparametrisation}).}
\label{tab-ESS-NUTS_SVI-1}
 \centering
\begin{tabular}{ccccccccccc}
\hline
    & $\Omega_b$ & $\Omega_c$ & $\sigma_8$ & $w_0$ & $h$ & $n_s$ & $A$ & $\eta$\\
\hline
% (0)  JEC extrait de https://github.com/DifferentiableUniverseInitiative/jax-cosmo-paper/blob/master/notebooks/VectorizedNumPyro_block.ipynb
%\color{red}(0) & $51.5$ &  $49.5$     & $39.3$     & $39.3$ & $52.0$ &  $50.2$ & $69.8$ & $46.2$\\
% (1) & (2) from JEC independant runs @ CCIN2P3
(1) & $48.1$ &  $45.6$     & $36.2$     & $33.4$ & $52.8$ & $50.1$ & $68.8$ & $48.8$\\
(2) & $38.7$ &  $24.8$     & $16.0$     & $18.6$ & $43.3$ & $39.5$ & $21.7$ & $16.6$\\
\hline
\end{tabular}
\end{table}

>>>>

The results of the evolution of the 2D parameter contours for some cosmological parameters compared to the default NUTS sampling (Sec.~\ref{sec-NUTS}) are displayed in figures \ref{fig_svi_nuts_1}, \ref{fig_svi_nuts_2}, \ref{fig_svi_nuts_3}. The agreement is quite good even with a rather low number of samples (ie. 200) as shown on figure \ref{fig_svi_nuts_1} and even for the highly correlated $(b_i)_{i:1,\dots,5}$ bias parameters (Fig.~\ref{fig_svi_nuts_4}). It is noticeable that running NUTS with the same settings  but without the SVI Neural Transport phase has demonstrated a worse behaviour.  Results in terms of sampling efficiency are shown on Tab.~\ref{tab-ESS-NUTS_SVI-1} showing that the SVI followed by Neutral Transport gives high efficiency like the default NUTS settings, without the need by the user to setup a particular mass matrix structuring (ie.\texttt{dense} matrix option). 
%
\subsection{Discussion}
%
The key question is what is the HMC/NUTS gain compared to the well known standard \texttt{Cobaya} sampler? 
In one hand, one may argue considering the elapse time of execution that the number of efficient samples per model call is a right metric to consider: 
\begin{equation}
    \eta = \frac{n_{eff}}{\#\ of\ calls} = \frac{N_s \times \varepsilon}{N_s \times n_{steps}} = \frac{\varepsilon}{n_{steps}}
\end{equation}
with $N_s$ the total number of samples, $\varepsilon$ the sampler efficiency and $n_{steps}$ the number of steps (calls) per sample generation. Then, for \texttt{cobaya} one roughly gets $\varepsilon\approx 1\%$ with $n_{steps}=1$ while for the NUTS sampler $\varepsilon\approx 50\%$ but at the expense of $n_{steps}=2^5$ or more. At the end of the day after tuning generation parameters, the $\eta$ values for both methods would certainly be comparable (ie. about $1\%$). So, at first sight one may conclude that there is no gain to use HMC/NUTS compared to standard MCMC sampler.

On the other hand, the above metric is may be too crude to cover all aspects of the sample generation. One should probably have in mind that the low sampling efficiency of standard MCMC sampler requires mobilizing a large amount of resources to produce a sufficient large sample batch in a reasonable time scale, ie. several days on dedicated infrastructure. Moreover, the user is faced to the dilemma to perform or not a statistical sample selection post-processing as \textit{thinning} beyond the \textit{burn-in removal} which is a rather delicate procedure and time consuming too, needing how-know to be conducted correctly \citep{doi:10.1146/annurev-statistics-040220-091727, 2015arXiv151007727O}.

So, the authors think that it is worth considering methods like HMC/NUTS that by construction are more efficient. Moreover, using SVI with neural reparametrisation offers an effective way to undertake a progressive validation of the model with rather modest sample set (eg. starting with 200 samples) producing good enough marginal contours in few hours. In practice, this validation phase can save time before producing sizeable batch for final analysis. The authors have not investigated very high dimensional ($O(10^2)$ parameters) as well as multi-modal analysis, but the key argument in favour of HMC/NUTS sampling is that it exploit the geometry of the typical set of the posterior distribution contrary to the standard random walk of Metropolis-Hasting sampling. Furthermore, using reparametrisation  one can adapt to poor geometry cases (eg. \cite{2019arXiv190303704H}).
%
\begin{table}[htb]
\caption{Results of the relative effective sample size (aka ESS, "bulk") in percent computed by the \texttt{Arviz} library \citep{arviz_2019}.
%{\color{red} 
%(0) "F.L run" are computed from NUTS sampler with 16 %parallel chains of 1,000 samples each and 200 warm-up %and \texttt{max\_tree\_depth=7} (see text) The numbers %are from \texttt{n\_eff/16,000};
%}
(1): results from NUTS sampler with 16 chains of 1,000 samples each and 200 warm-up and \texttt{max\_tree\_depth=7} (Sec.~\ref{sec-NUTS}); (2) are the results of SVI Multivariate Normal followed by NUTS and Neural Transform  with 10 chains of 1,000 samples each and 200 warm-up (Sec.~\ref{sec-Neural-Reparametrisation}).}
\label{tab-ESS-NUTS_SVI-1}
 \centering
\begin{tabular}{ccccccccccc}
\hline
    & $\Omega_b$ & $\Omega_c$ & $\sigma_8$ & $w_0$ & $h$ & $n_s$ & $A$ & $\eta$\\
\hline
% (0)  JEC extrait de https://github.com/DifferentiableUniverseInitiative/jax-cosmo-paper/blob/master/notebooks/VectorizedNumPyro_block.ipynb
%\color{red}(0) & $51.5$ &  $49.5$     & $39.3$     & $39.3$ & $52.0$ &  $50.2$ & $69.8$ & $46.2$\\
% (1) & (2) from JEC independant runs @ CCIN2P3
(1) & $48.1$ &  $45.6$     & $36.2$     & $33.4$ & $52.8$ & $50.1$ & $68.8$ & $48.8$\\
(2) & $38.7$ &  $24.8$     & $16.0$     & $18.6$ & $43.3$ & $39.5$ & $21.7$ & $16.6$\\
\hline
\end{tabular}
\end{table}
